---
title: "Cross-validation with rgs3"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
colorlinks: true
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Cross-validation with rgs3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
setwd("~/src/rgs3/vignettes/")

library(devtools)
build_vignettes()

library(rmarkdown)
render("crossval-rgs3.Rmd", "html_document")
-->

```{r knitr_setup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.align="center")
```


# Preamble

Load the `rgs3` package (available [here](https://github.com/INRA/rgs3)):
```{r load_pkg}
suppressPackageStartupMessages(library(rgs3))
packageVersion("rgs3")
```

Choose the number of cores to run cross validation in parallel (if available):
```{r}
suppressPackageStartupMessages(library(parallel))
nb.cores <- max(detectCores() - 1, 1)
```

Then, read first the vignette `intro-rgs3` before this one.

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Data and model

This vignette uses the same data as the `intro-rgs3` vignette.



# Load data into R

Load phenotypes:

```{r load_phenos}
phenos.file <- system.file("extdata", "phenos_df.txt.gz", package="rgs3")
tools::md5sum(path.expand(phenos.file))
phenos <- read.table(phenos.file, header=TRUE)
phenos$year <- as.factor(phenos$year)
str(phenos)
```

Load genotypes:

```{r load_genos}
genos.file <- system.file("extdata", "genos_mat.txt.gz", package="rgs3")
tools::md5sum(path.expand(genos.file))
genos <- as.matrix(read.table(genos.file))
dim(genos)
```



# Set up the configuration

Let us choose an identifier for our analysis, which will be used for all files, so that it will be easy to remove them at the end:

```{r task_id}
task.id <- "task-execGS3_crossval-rgs3"
```

Use the utility function:

```{r config_from_default}
ptl <- data.frame(position = c(which(colnames(phenos) == "year"),
                               ncol(phenos) + 1),
                  type = c("cross",
                           "add_SNP"),
                  nlevels = c(length(levels(phenos$year)),
                              0),
                  stringsAsFactors = FALSE)
config <- getDefaultConfig(
    nb.snps = ncol(genos),
    rec.id = which(colnames(phenos) == "geno"),
    twc = c(which(colnames(phenos) == "response1"), 0),
    method = "VCE",
    ptl = ptl,
    task.id = task.id)
```

Customize the rest:

```{r config_rest}
vc <- data.frame(var = c("vara","vard","varg","varp", "vare"),
                 exp = c("0.025", "0", "0", "0", "1"),
                 df = rep("2", 5),
                 stringsAsFactors = FALSE)
config$vc <- vc
config$niter <- 2000
config$burnin <- 200
config$thin <- 2
```

Note that the number of iterations here is voluntarily low, so that the vignette doesn't take too long to run.


# Perform K-fold cross-validation

Let's use 5 folds for this vignette:

```{r cv}
results.cv <- crossValWithGs3(
    genos = genos,
    dat = phenos,
    config = config,
    task.id = task.id,
    ped.file = NULL,
    nb.folds = 5,
    seed = 123,
    remove.files = "all",
    nb.cores = nb.cores,
    verbose = 2)
```


# Assess the results

```{r}
results.cv
apply(results.cv[,c("rmspe","cor.p","cor.p.div.h","reg.intercept","reg.slope")],
      2, function(x){
        c(mean=mean(x), sd=sd(x))
      })
```

It is also possible to have multiple replicates, using the function `crossValRepWithGs3`.



# Acknowledgments

* Hélène Muranty


# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
