---
title: "Cross-validation with rgs3"
author: "Hélène Muranty and Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: TRUE
  html_document:
    toc: true
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
colorlinks: true
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Cross-validation with rgs3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
setwd("~/src/rgs3/vignettes/")

library(devtools)
build_vignettes()

library(rmarkdown)
render("crossval-rgs3.Rmd", "html_document")
-->

```{r knitr_setup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.align="center")
```


# Preamble

Load the `rgs3` package (available [here](https://github.com/timflutre/rgs3)):
```{r load_pkg}
library(rgs3)
packageVersion("rgs3")
```

Then, read first the vignette `intro-rgs3` before trying to run this one.

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Data and model

This vignette uses the same data as the `intro-rgs3` vignette.



# Load data into R

Load phenotypes:

```{r load_phenos}
phenos.file <- system.file("extdata", "phenos_df.txt.gz", package="rgs3")
phenos <- read.table(phenos.file, header=TRUE)
phenos$year <- as.factor(phenos$year)
str(phenos)
```

Load genotypes:

```{r load_genos}
genos.file <- system.file("extdata", "genos_mat.txt.gz", package="rgs3")
genos <- as.matrix(read.table(genos.file))
dim(genos)
```



# Set up the configuration

Use the utility function:

```{r config_from_default}
ptl <- data.frame(position = c(which(colnames(phenos) == "year"),
                               ncol(phenos) + 1),
                  type = c("cross",
                           "add_SNP"),
                  nlevels = c(length(levels(phenos$year)),
                              0),
                  stringsAsFactors = FALSE)
config <- getDefaultConfig(
    nb.snps = ncol(genos),
    rec.id = which(colnames(phenos) == "geno"),
    twc = c(which(colnames(phenos) == "response1"), 0),
    method = "VCE",
    ptl = ptl)
```

Customize the rest:

```{r config_rest}
vc <- data.frame(var = c("vara","vard","varg","varp", "vare"),
                 exp = c("0.025", "0", "0", "0", "1"),
                 df = rep("2", 5),
                 stringsAsFactors = FALSE)
config$vc <- vc
config$niter <- 2000
config$burnin <- 200
config$thin <- 2
```

Note that the number of iterations here is voluntarily low, so that the vignette doesn't take too long to run.


# Perform K-fold cross-validation

Let's use 5 folds for this vignette:

```{r cv}
results.cv <- crossValWithGs3(
    genos = genos,
    dat = phenos,
    ped.file = NULL,
    nb.folds = 5,
    config = config,
    seed = 123,
    remove.files = "all", # use "some" in real-life applications
    verbose = 2)
```


# Assess the results

```{r}
results.cv
```



# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
